# RStudio basics

At this point you should have an operational RStudio terminal at your disposal.  If not, go back to [Getting started].  This chapter will focus some of the most basic tools and skills you'll need to successfully use RStudio. We'll start with using R as a calculator, learn about naming and storing values/objects, then progress to loading and using data frames (the bread and butter of R in some sense), and finish with R Markdown documents, which allow you to create fully reproducible documents that contain a mixture of text and code.  Fun fact: these notes were written in a collection of RMarkdown documents and compiled with a package called `bookdown`.  

As you proceed, we encourage you to try to run any code you see in this book and to play around and do things on your own. The best way to learn how to code is to experiment. The immediate feedback of running some successful code can be quite gratifying!

## A big calculator

At its most basic, R is a big, fancy, and initially cumbersome calculator. Think of any calculation you'd like to do and R can likely do it. R can handle more than just numbers!  For our purposes, we'll need numbers, strings, and logicals.   

### Numbers

R has three "atomic" types of numbers: real, integer, and complex, but we will only need to think of real numbers.  You can do arithmetic and evaluate the standard elementary functions with numbers as you would expect.

- Addition, multiplication, subtraction, and division are `+`, `*`, `-`, and `/` respectively. For example: 

```{r}
2 + 2
3 * 1.5
9 - 4
5/3
```

- Exponentiation is either `^` or `**`, the latter is old school. You can raise a number to a power of 10 using scientific notation, for example `1.2e4` is 12000. 

```{r}
2^3
2**3
1.2e4
```
- All of your favorite "elementary" functions are built into R. Note that the `log` function is base $e$, not base 10.  The example below shows how to change base.

```{r}
exp(2) # e^2
sin(pi/2) # note the trig functions are in radians
log(10) # natural log of 10
log(100, 10) # the log of 100, base 10
```
**Reminder!** Be careful and mindful about the order of operations; computers and calculators read your input very literally. For instance, $6 \div 2(1 + 2)$, is an annoying statement with respect to the order of operations. R will evaluate this expression correctly if you type `6/2*(1+2)`. Make sure you can evaluate it correctly as well so that you input what you mean.

### Strings

We want to be able to analyze both numerical and categorical variables in R.  For this reason, along with many others, your can use and manipulate strings of character in R. We won't be doing much more than using strings for values of categorical variables, so we won't go into too much detail. 

To define a string, simply put the expression in question in quotation marks.

```{r, eval=F}
"R might be cool? I don't know."
```

Base R has some handy functions for manipulating strings. We may not need all of them, but they're good to be aware of. 

- `substr(string, start=n1, stop=n2)` will return a subset of your string starting at the `n1`th character, ending at the `n2` character. 

```{r}
substr("R might be cool? I don't know.", start = 18, stop = 30)
```

- `nchar(string)` counts the number of characters in a string. 

```{r}
nchar("R might be cool? I don't know.")
```

- `toupper/tolower(string)` converts all letters to either upper or lower case. 

```{r}
toupper("R might be cool? I don't know.")
```

- `paste(..., sep = " ")` will concatenate a collection of strings, separated by a space. You can change what separates the strings. This function is most useful when you have to repeatedly concatenate some strings.  

```{r}
paste("R might be cool? I don't know.", "Yes, it is!")
```

You can do a lot more with strings, but things start to get more complicated quickly.  The package `stringr` has some really handy functions, but are unnecessary for this course. 

### Logicals and Logical operators

Throughout this class you will need to compare various objects in R using standard "logical operators" like "equals" (`==`), "less than" `<`, "greater than or equal to `>=`" etc. When you compare objects using these operators, R returns a new type of object called a "logical". Logicals are just `TRUE` and `FALSE`. 

- You can check equality of numbers and strings using `==`. To check if two objects are unequal, use `!=`.

```{r}
4 == 5
2^3 == 8
"cat" == "dog"
"dog" != "cat"
```

- You can compare numbers with your favorite inequalities (`<`, `>`, `<=`, `>=`) as well. More interestingly, you can also compare strings with inequalities. R defaults to lexicographic (ie dictionary) ordering of strings.

```{r}
10 < 9
exp(3) >= exp(2)
"a" < "b"
"aa" > "ab"

```

- For convenience, `TRUE` sometimes behaves like 1 and `FALSE` like 0. In particular, you can perform arithmetic with logicals.  This makes counting the number of matches you have quite easy in some situations, as we'll see.

```{r}
TRUE + TRUE
TRUE*FALSE
```

- Note: R may occasionally exhibit behavior that you find goofy when comparing objects. This is usually because of *coercion*.  When comparing objects, R will try to make the objects into the same type, if possible. We don't want to get bogged down in these details, but it is something to be aware of. The following example illustrates coercion when trying to compare a string to a number. 

```{r}
"1" == 1 #R converts the string to 1, as a number, then compares.
"01" == 1 # R leaves the string as is, so the objects are unequal.
```

## Naming things

Now that we know what types of objects we can use in R, we can talk about naming objects and storing them in your computers memory so that you can reference them easily later. You can name objects in two ways, either with the classic `<-` or with `=`. These two methods of naming objects technically do different things, but the difference is subtle enough that we won't worry about it. Note that when you name an object, it will appear in your "Environment" window in RStudio. The example below will illustrates the basics of naming objects. 

```{r}
x <- 1.123e6
y <- 5
x 
y
x/y 
```

**Technical aside. Only read this if you have experience programming elsewhere, skip to next section otherwise.** R makes copies of objects in a different way from, say, Python (and many other languages).The difference lies in deep vs shallow copies. Basically, a deep copy creates an entirely new object in memory, but a shallow copy merely points to the original object's position in memory. R will actually create a shallow copy until the original object is modified, then it will create a deep copy in order to save on memory. To illustrate this point, if you were to run the following code in Python

```{r, eval = FALSE}
x = [1,2,3]
y = x #shallow copy created here
x.append(5)
y
```

it would print `[1,2,3,5]` since, in Python, `y` was a shallow copy of `x`.  On the other had, the equivalent code in R is

```{r}
x = 1:3
y = x # shallow copy of x created here
x <- append(x, 5) #but deep copy of original x created here
y
```
It's good to know how R is behaving; deep copies can end up using more memory, but that won't be an issue in this class with modern computing power

## Lists and vectors

### Creating lists

### Subsetting lists

## Packages

## Data frames and tibbles

### Whats the difference? 

### Using data frames/tibbles

### Creating data frames/tibbles

#### Manual entry

#### Reading files

## R Markdown documents 
