# RStudio basics

At this point you should have an operational RStudio terminal at your disposal.  If not, go back to [Getting started].  This chapter will focus some of the most basic tools and skills you'll need to successfully use RStudio. We'll start with using R as a calculator, learn about naming and storing values/objects, then progress to loading and using data frames (the bread and butter of R in some sense), and finish with R Markdown documents, which allow you to create fully reproducible documents that contain a mixture of text and code.  Fun fact: these notes were written in a collection of RMarkdown documents and compiled with a package called `bookdown`.  

As you proceed, we encourage you to try to run any code you see in this book and to play around and do things on your own. The best way to learn how to code is to experiment. The immediate feedback of running some successful code can be quite gratifying!

## A big calculator

At its most basic, R is a big, fancy, and initially cumbersome calculator. Think of any calculation you'd like to do and R can likely do it. R can handle more than just numbers!  For our purposes, we'll need numbers, strings, and logicals.   

### Numbers

R has three "atomic" types of numbers: real, integer, and complex, but we will only need to think of real numbers.  You can do arithmetic and evaluate the standard elementary functions with numbers as you would expect.

- Addition, multiplication, subtraction, and division are `+`, `*`, `-`, and `/` respectively. For example: 

```{r}
2 + 2
3 * 1.5
9 - 4
5/3
```

- Exponentiation is either `^` or `**`, the latter is old school. You can raise a number to a power of 10 using scientific notation, for example `1.2e4` is 12000. 

```{r}
2^3
2**3
1.2e4
```
- All of your favorite "elementary" functions are built into R. Note that the `log` function is base $e$, not base 10.  The example below shows how to change base.

```{r}
exp(2) # e^2
sin(pi/2) # note the trig functions are in radians
log(10) # natural log of 10
log(100, 10) # the log of 100, base 10
```
**Reminder!** Be careful and mindful about the order of operations; computers and calculators read your input very literally. For instance, $6 \div 2(1 + 2)$, is an annoying statement with respect to the order of operations. R will evaluate this expression correctly if you type `6/2*(1+2)`. Make sure you can evaluate it correctly as well so that you input what you mean.

### Strings

We want to be able to analyze both numerical and categorical variables in R.  For this reason, along with many others, your can use and manipulate strings of character in R. We won't be doing much more than using strings for values of categorical variables, so we won't go into too much detail. 

To define a string, simply put the expression in question in quotation marks.

```{r, eval=F}
"R might be cool? I don't know."
```

Base R has some handy functions for manipulating strings. We may not need all of them, but they're good to be aware of. 

- `substr(string, start=n1, stop=n2)` will return a subset of your string starting at the `n1`th character, ending at the `n2` character. 

```{r}
substr("R might be cool? I don't know.", start = 18, stop = 30)
```

- `nchar(string)` counts the number of characters in a string. 

```{r}
nchar("R might be cool? I don't know.")
```

- `toupper/tolower(string)` converts all letters to either upper or lower case. 

```{r}
toupper("R might be cool? I don't know.")
```

- `paste(..., sep = " ")` will concatenate a collection of strings, separated by a space. You can change what separates the strings. This function is most useful when you have to repeatedly concatenate some strings.  

```{r}
paste("R might be cool? I don't know.", "Yes, it is!")
```

You can do a lot more with strings, but things start to get more complicated quickly.  The package `stringr` has some really handy functions, but are unnecessary for this course. 

### Logicals and Logical operators

Throughout this class you will need to compare various objects in R using standard "logical operators" like "equals" (`==`), "less than" `<`, "greater than or equal to `>=`" etc. When you compare objects using these operators, R returns a new type of object called a "logical". Logicals are just `TRUE` and `FALSE`. 

- You can check equality of numbers and strings using `==`. To check if two objects are unequal, use `!=`.

```{r}
4 == 5
2^3 == 8
"cat" == "dog"
"dog" != "cat"
```

- You can compare numbers with your favorite inequalities (`<`, `>`, `<=`, `>=`) as well. More interestingly, you can also compare strings with inequalities. R defaults to lexicographic (ie dictionary) ordering of strings.

```{r}
10 < 9
exp(3) >= exp(2)
"a" < "b"
"aa" > "ab"

```

- For convenience, `TRUE` sometimes behaves like 1 and `FALSE` like 0. In particular, you can perform arithmetic with logicals.  This makes counting the number of matches you have quite easy in some situations, as we'll see.

```{r}
TRUE + TRUE
TRUE*FALSE
```

- Note: R may occasionally exhibit behavior that you find goofy when comparing objects. This is usually because of *coercion*.  When comparing objects, R will try to make the objects into the same type, if possible. We don't want to get bogged down in these details, but it is something to be aware of. The following example illustrates coercion when trying to compare a string to a number. 

```{r}
"1" == 1 #R converts the string to 1, as a number, then compares.
"01" == 1 # R leaves the string as is, so the objects are unequal.
```

## Naming things

Now that we know what types of objects we can use in R, we can talk about naming objects and storing them in your computers memory so that you can reference them easily later. You can name objects in two ways, either with the classic `<-` or with `=`. These two methods of naming objects technically do different things, but the difference is subtle enough that we won't worry about it. Note that when you name an object, it will appear in your "Environment" window in RStudio. The example below will illustrates the basics of naming objects. 

```{r}
x <- 1.123e6
y <- 5
x 
y
x/y 
```

Once you've named an object and stored it in memory, you can use its name anywhere else to call the value associated to the name. 

**Read the following only if you've had some programming experience. If not, skip to the next section.** R makes copies of objects in a different way from, say, Python (and many other languages).The difference lies in deep vs shallow copies. Basically, a deep copy creates an entirely new object in memory, but a shallow copy merely points to the original object’s position in memory. R will actually create a shallow copy until the original object is modified, then it will create a deep copy in order to save on memory. To illustrate this point, if you were to run the following code in Python:

```{r, eval = FALSE}
x = [1,2,3]
y = x #shallow copy created here
x.append(5)
y
```

it would print `[1,2,3,5]`. However, the equivalent code in R is

```{r}
x = 1:3
y = x # shallow copy of x created here
x <- append(x, 5) #but deep copy of original x created here for y
y
```

It’s good to know how R is behaving; deep copies can end up using more memory, but that won’t be an issue in this class with modern computing power.

## Lists and vectors

One of the many advantages to using a computer for mathematics, statistics, and data analysis is their ability to store and organize large chunks of information. A list/vector is the most basic way to start organizing data in R. This section will guide you through creating and manipulating lists. 

### Creating lists

The most basic tool for creating a list in R is the function `c(..)`.  According to the documentation `c(...)` "is a generic function which combines its arguments."  All that to say that it is a function that creates a list. Here's a few examples. You can make lists with any types of data in them. 

```{r}
x <- c(1,2,3)
y <- c("cat", "dog", "rabbit")
```

You can combine lists easily with the `c(...)` function as well, but notice in the example below that something funny happens.

```{r}
z <- c(x, y)
z
```

R coerced the numbers from the list `x` into characters/strings because `y` contains strings. This is a subtlety that can sometimes cause headaches.  If your lists all contain the same types of objects, you have nothing to worry about. 

Throughout this course you'll often want to know how many objects are in a list and most times we don't want to manually count them. The `length` function does this for us. The examples above are easy to count by hand, but as a quick check:

```{r}
length(x)
length(y)
length(z)
```

### Subsetting lists

After defining a list, sometimes we want to access only certain elements of that list; this is called **subsetting**. We'll spend more time on fancier subsetting methods later in this course, but this will get us started with the basics. *Technical note:* R indexing starts at 1 and not 0, a difference between R and some other programming languages like python. 

To get started, we'll create a long list of randomly generated numbers. 

```{r}
rando <- runif(1000, min = 0, max = 100)
length(rando)
```

This list has 1000 elements in it. We can find the 497th element using

```{r}
rando[497]
```

and in general, use `listName[number]`. We can also easily select ranges of indicies. For instance, suppose we want to know what the first 10 values in our list are.  

```{r}
rando[1:10]
```

Finally, we can pick out specific elements from our big list using a smaller list of our desired indices. For example, if we wanted the entries number 2,3,5,7,11, and 13 from our list, we could use the following.

```{r}
indicies <- c(2,3,5,7,11,13)
rando[indicies]
```


### Special lists

The last section taught us how to make the most basic type of list. While useful, making lists in this way can be tedious because it requires one to type in all values. Often out in the wild, you will want to create lists that follow specific patterns. For instance: the letters "c" through "p"; a list of the integers from 1 to 100; a list of numbers from -10 to 10, spaced by .01; the letters "d", "g", and "z" repeated 1000 times each. It is easy to construct these types of lists in R with minimal typing. 

*Lists of integers*: to make a list of sequential integers in R, use the format `start : end`, for instance `2:5` will return the list `2,3,4,5`.

```{r}
x <- -8:12
x
length(x)
```

*Lists of sequential numbers that are not integers:* The function `seq` produces sequences of numbers that follow a particular pattern. The general syntax is `seq( from = STARTING VALUE, to = ENDING VALUE, by = SPACE BETWEEN ENTRIES)`

So if we wanted a list of all numbers from -10 to 10, spaced by .01, we'd use

```{r}
foo <- seq(from = -10, to = 10, by = .01)
```

We can access the first 5 elements of this list and count the number of elements in our list `foo` with the following.  

```{r}
foo[1:5]
length(foo)
```

*Lists of sequential letters:* R has two in-built lists of letters called `letters` and `LETTERS`. No surprise, one of them is a list of lower case letters, the other upper case. 

```{r}
letters
```

To select the 4th through 11th lowercase letters of the alphabet, use the following syntax (which will be explained in a subsequent section). 

```{r}
letters[4:11]
```

*Lists with a repeating patterns:* 



### List arithmetic

## Packages

## Data frames and tibbles

### Whats the difference? 

### Using data frames/tibbles

### Creating data frames/tibbles

#### Manual entry

#### Reading files

## R Markdown documents 
